# Introduction {#intro}

You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. This sample book was built on top of R Markdown and **knitr** [@xie2015].

## Problem description

<!--- Rewrite this for problem identification.

The mailing included a gift (or "premium") of personalized name &
address labels plus an assortment of 10 note cards and envelopes.  All
of the donors who received this mailing were acquired by PVA through
similar premium-oriented appeals such as this.

One group that is of particular interest to PVA is "Lapsed" donors.
These are individuals who made their last donation to PVA 13 to 24
months ago.  They represent an important group to PVA, since the
longer someone goes without donating, the less likely they will be to
give again.  Therefore, recapture of these former donors is a critical
aspect of PVA's fund raising efforts.

However, PVA has found that there is often an inverse correlation
between likelihood to respond and the dollar amount of the gift, so a
straight response model (a classification or discrimination task) will
most likely net only very low dollar donors.  High dollar donors will
fall into the lower deciles, which would most likely be suppressed
from future mailings.  The lost revenue of these suppressed donors
would then offset any gains due to the increased response rate of the
low dollar donors.

Therefore, to improve the cost-effectiveness of future direct
marketing efforts, PVA wishes to develop a model that will help them
maximize the net revenue (a regression or estimation task) generated
from future renewal mailings to Lapsed donors.
-->
