--- 
title: "Topic"
author: "Florian Hochstrasser"
date: "`r Sys.Date()`"
output: pdf_document
description: Thesis for the programme Master in Statistics at the University of Neuchâtel,
  Switzerland.
documentclass: article
link-citations: yes
bibliography: book.bib
site: bookdown::bookdown_site
subtitle: Master Thesis
biblio-style: apalike
---

# Abstract {-}

It is known that....

This work addressed the problem from the perspective of...

It could be shown that...

Henceforth, it should be considered that...

<!--chapter:end:index.Rmd-->

# Introduction {#intro}

You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. This sample book was built on top of R Markdown and **knitr** [@xie2015].

<!--chapter:end:01-intro.Rmd-->

# Data

The data at hand originates from a U.S. veterans organisation and is freely available online[^1]. It contains data from the organisations members and the turnout of a direct mailing addressed to 3.5 million members in the scope of a fundraising effort that was conducted in 1997.
[1]: See [UCI Machine Learning Repository: KDD Cup 1998 Data](https://archive.ics.uci.edu/ml/datasets/KDD+Cup+1998+Data)

In the following section, the data will be characterized.




<!--chapter:end:02-data.Rmd-->

# Methods

We describe our methods in this chapter.

<!--chapter:end:03-method.Rmd-->

# Applications

Some _significant_ applications are demonstrated in this chapter.

## Example one

## Example two

<!--chapter:end:04-application.Rmd-->

# Final Words

We have finished a nice book.

<!--chapter:end:05-summary.Rmd-->

`r if (knitr::is_html_output()) '
# References {-}
'`

<!--chapter:end:06-references.Rmd-->

