\begin{titlepage}

\titlehead{Master Thesis}
\title{Profit maximization for direct marketing campaigns}
\subject{An application of data-driven decision making}
\subtitle{Submitted in partial fulfillment of the requirements for the degree of Master of Science in Statistics}
\author{Florian Hochstrasser}
\date{14.06.2019}
\publishers{Supervisor: Jacques Zuber}

\end{titlepage}

\let\maketitle\oldmaketitle
\maketitle

\begin{abstract}
A profit optimization problem, based on the data distributed for the KDD-Cup 1998, was solved using several different supervised learning algorithms. A complete analysis including data preprocessing, model evaluation and selection and construction of the final estimator was conducted.

During preprocessing, several approaches for imputation and feature selection were explored. A two-stage prediction approach was chosen, separately predicting donation probability and donation amount. Five classification- and four regression-algorithms were evaluated. The best-performing algorithms of both families were then combined for the final prediction.

A python package was developed to automate preprocessing of the data, guaranteeing consistent treatment of learning- and validation data. The package features an extensible architecture for future work on the data set.

In comparison to the winners of the Cup, the developed model ranked $4^{th}$. Future work based on the results achieved should focus on feature engineering and alternative imputation strategies.
\end{abstract}
